---
title: "Table of Links to Reports"
date: '`r format(as.POSIXct(if ("" != Sys.getenv("SOURCE_DATE_EPOCH")) { as.numeric(Sys.getenv("SOURCE_DATE_EPOCH")) } else { Sys.time() }, origin="1970-01-01"), "%Y-%m-%d %H:%M:%S")`'
params:
  sample_sheet: ''
---


```{r, child="_settings.Rmd"}
#Read config.yml for the locations of sample_sheet 
```

# Overview of Reports

For each sample that was provided three report pages were generated. These can be found here by clicking on the provided
link or by navigating through the tabs in the menu bar. 

```{r read sample file, echo=F}

sample_sheet <- fread(params$sample_sheet)
sample_names <- sample_sheet$name
report_name <- c("variant_report", "qc_report","taxonomic_classification")
report_link <- c(".variantreport_p_sample.html",".qc_report_per_sample.html",".taxonomic_classification.html")

```


```{r links, echo=FALSE}
#todo: automatically generate the sample links
#todo: make a table for this overview
for (sample in sample_names){
  print(sample)
}

```

* __BERLIN__
  + sample 1
    - variants report
    - qc report
    - taxonomic classification
  + sample 2
    - variants report
    - qc report
    - taxonomic classification

* __Munich__
  + sample 3
    - variants report
    - qc report
    - taxonomic classification
  + sample 4
    - variants report
    - qc report
    - taxonomic classification