---
title: "index page"
date: '`r format(as.POSIXct(if ("" != Sys.getenv("SOURCE_DATE_EPOCH")) { as.numeric(Sys.getenv("SOURCE_DATE_EPOCH")) } else { Sys.time() }, origin="1970-01-01"), "%Y-%m-%d %H:%M:%S")`'
output:
  html_document:
    theme: lumen
    highlight: monochrome
    code_folding: hide
    depth: '3'
    toc: yes
    toc_float: yes
    number_sections: yes
params:
  sample_sheet: ''
---




# Overview of Reports

For each sample that was provided three report pages were generated. These can be found here by clicking on the provided link or by navigating through the tabs of this website.

```{r read sample file, echo=F}
library(data.table)
sample_sheet <- fread(params$sample_sheet)
sample_names <- sample_sheet$names
```


```{r links, echo=FALSE}
#todo: automatically generate the sample links
#todo: make a table for this overview
for (sample in sample_names){
  print(sample)
}

```

* __BERLIN__
  + sample 1
    - variants report
    - qc report
    - taxonomic classification
  + sample 2
    - variants report
    - qc report
    - taxonomic classification

* __Munich__
  + sample 3
    - variants report
    - qc report
    - taxonomic classification
  + sample 4
    - variants report
    - qc report
    - taxonomic classification