---
title: "SARS-CoV-2 QC report"
date: '`r format(as.POSIXct(if ("" != Sys.getenv("SOURCE_DATE_EPOCH")) { as.numeric(Sys.getenv("SOURCE_DATE_EPOCH")) } else { Sys.time() }, origin="1970-01-01"), "%Y-%m-%d %H:%M:%S")`'
output:
  html_document:
    theme: lumen
    highlight: monochrome
    code_folding: hide
    depth: '3'
    toc: yes
    toc_float: yes
    number_sections: yes
params:
  sample_name: ''
  coverage_file: ''
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(dplyr)))
```

# Input Settings

### Sample: `r params$sample_name`
```{r reading data, include = FALSE}
coverage.df <- read.table(params$coverage_file, sep = "\t", header = T)
coverage.df <- na_if(coverage.df, "[]")
coverage.df <- gsub('\\[','',coverage.df)
coverage.df <- gsub('\\]','',coverage.df)
# fixme: this can be sure done in a nicer and more compact way
```
## Amplicon coverage  
The information about the amplicon primer was taken from the [ARTIC bioinformatics platform](https://github.com/joshquick/artic-ncov2019/blob/master/primer_schemes/nCoV-2019/V3/nCoV-2019.bed).
The reference name "MN908947.3" was changed to "NC_045512.2". Both are describing the same reference genome for 
[SARS-CoV-2](https://www.ncbi.nlm.nih.gov/nuccore/1798174254). 
 
```{r, message=F, warning=F, echo = FALSE}

AmpliconCoverage <- data.frame("Number of amplicon fully covered" = paste0(coverage.df[1],' out of 98'), 
                  "Anmplicons partially covered" = coverage.df[2],
                  "Drop-out amplicons" = coverage.df[3]) # maybe more clear title stating which coverage

DT::datatable(t(AmpliconCoverage), options = list(searching = FALSE, paging = FALSE))

```

## Coverage

```{r, message=F, warning=F, echo = FALSE}

Coverage <- data.frame("Total coverage" = coverage.df[4], 
                  "Aligned reads" = paste(round(as.integer(coverage.df[5]),1),"%"),
                  "Mean depth" = round(as.integer(coverage.df[6]))) # maybe more clear title stating which coverage

DT::datatable(t(Coverage), options = list(searching = FALSE, paging = FALSE))

```
## FastQC and Trimming

Download: [fastqc report of unprocessed data](/path/to/fastqc.html)  
Download: [fastqc report of pre-processes data](/path/to/fastqc.html) 



The reads were pre-processed as followed:  

|      |     |  
|:-----|:----|  
|trim length cutoff prinseq |0.2|

## Download Stats csv
```{r, include = F}

write.csv2(coverage.df, file.path(dirname(params$coverage_file), "report_download_coverage.csv")) # not sure about the funct of the dataframe here, but it works so I leave it for now

readLines(params$coverage_file) %>%
  paste0(collapse="\n") %>% 
  openssl::base64_encode() -> coverage
```
Coverage statistics:  
[Download Coverage_stats.csv](`r sprintf('data:text/csv;base64,%s', coverage)`)

