---
title: "PiGx: BS-seq"
output: 
  html_notebook:
    toc:        TRUE
    toc_float:  TRUE
    theme:      "lumen"
    number_sections: FALSE
    code_folding: "hide"
    self_contained: TRUE
    includes:
      in_header: pigx_bsseq_logo.html
date: "`r Sys.Date()`"
#bibliography: reports.bib
params:
  inputfiles: ["methylation_calls/sampleA_1_val_1_bt2.deduped.sorted_CpG.txt","methylation_calls/sampleBse_se_bt2.deduped.sorted_CpG.txt"]
  methylDiff_file: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RDS"
  methylDiff_hyper_file: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RDS"
  methylDiff_hypo_file: "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows_gr.RDS"
  outBed:   "segmentation/sampleA.pe1_trimmed_bismark_bt2.deduplicated.sorted_meth_windows.bed"
  sampleids: ["sampleA", "sampleB"]
  assembly: "ce10"
  treatment: [0,1]
  mincov: 0
  context: "CpG"
  workdir: "~/outputdir/"
  cores: 2
  sessioninfo: TRUE
  references: TRUE
---


```{r include=FALSE}
makeSessioninfo=params$sessioninfo
makeReferences = params$references
```
  
  
```{r eval_params, echo=FALSE}

#saveRDS(params, "~/params.RDS")
#params = readRDS("~/params.RDS")
#print(getwd())

input     <- params$inputfiles
sampleids <- params$sampleids
output    <- params$outBed
methylDiff_file   <- params$methylDiff_file
methylDiff_hyper_file    <- params$methylDiff_hyper_file
methylDiff_hypo_file   <- params$methylDiff_hypo_file
assembly  <- params$assembly
treatment <- as.numeric(params$treatment)
mincov    <- as.numeric(params$mincov)
context   <- params$context
workdir   <- params$workdir
cores     <- as.numeric(params$cores)


```


## Differential methylation report

### Calling differentially methylated bases

The logistic regression is applied to model the log odds ratio which is based on methylation proportion of a CpG using the treatment vector which denotes the sample group membership for the CpGs in the model. 

After q-value calculation, differentially methylated bases are extracted based on q-value and percent methylation difference cutoffs. Here we select bases that have q-value < 0.01 and percent methylation difference larger than 25%. Futhermore, we calculate hyper-methylated or hypo-methylated bases. 

Overdispersion occurs when there is more variability in the data than assumed by the distribution and is here included in the differentially methylatation calculation.

For more details about `calculateDiffMeth()` and `getMethylDiff()` functions see [@methylKit2012] and for details about
the logistic regression and overdispersion see [@Wreczycka2017].


```{r, include=FALSE,warning=FALSE}

suppressPackageStartupMessages(expr = {
  ## load methylKit
  library("methylKit",warn.conflicts = FALSE,quietly = TRUE)
  library("DT",warn.conflicts = FALSE,quietly = TRUE)
})

```

```{r}
# This code is to test if diff. meth reports will work on data that contain diff. meth. cytosines.
# by using fake input files (takes from the methylKit library).
#fake_input=c(  "/home/kwreczy/repositories/pigx_bsseq/test_CElegans/tmp/test1.myCpG1.txt",
#              "/home/kwreczy/repositories/pigx_bsseq/test_CElegans/tmp/control2.myCpG1.txt"
#              )
#file.copy(fake_input[1], paste0(workdir, input)[1], overwrite=TRUE)
#file.copy(fake_input[2], paste0(workdir, input)[2], overwrite=TRUE)

```

### Libraries
```{r load libraries, eval=FALSE, include=TRUE}

## load libraries
library("methylKit")
library("DT")

```

### Parameters

**Input parameters:**
```{r print_params_in, echo=TRUE}

inputParams <- data.frame(Sample.id=sampleids,
                          Input = input,
                          Treatment=treatment,
                          Context=context,
                          Mincov=mincov,
                          Assembly=assembly,
                          stringsAsFactors = FALSE)
inputParams.t = t(inputParams)
colnames(inputParams.t) = c("Sample1", "Sample2")

knitr::kable(inputParams.t)


```

**Output parameters:**
```{r print_params_out, echo=TRUE}

paramsDesc <- c("BEDfile", "RDSfile (diff. methyl. Cs)","RDSfile_hyper (hyper-methyl. Cs)","RDSfile_hypo (hypo-methyl. Cs)")
paramsValues <- c(output, methylDiff_file, methylDiff_hyper_file, methylDiff_hypo_file)
outputParams <- data.frame(Parameters = paramsDesc,
                          Values = paramsValues,
                          stringsAsFactors = FALSE)
knitr::kable(outputParams)

```


### Find differentially methylated cytosines

```{r load methylation, echo=TRUE, warning=FALSE}

# Load methylation profiles

# Read input files
inputfiles = paste0(workdir, input)
methRawList.obj = methRead(as.list(inputfiles),
                                sample.id = as.list(sampleids),
                                assembly = assembly,
                                treatment=treatment,
                                context = context,
                                mincov = mincov
)

# Take bases with coverage in all samples.
meth.unite=unite(methRawList.obj, destrand=FALSE)

meth.unite.nonempty = nrow(meth.unite)>1
if(!meth.unite.nonempty) print("There are no bases with coverage in all samples")


# Cutoff for absolute value of methylation percentage change
# between test and control
difference = 25
# Cutoff for qvalue of differential methylation statistic
qvalue = 0.01



if(nrow(meth.unite)>1){

      meth.diffmeth <- calculateDiffMeth(meth.unite, 
                                   overdispersion="MN",
                                   adjust = "qvalue",
                                   test="Chisq",
                                   mc.cores=cores) 

      # Get differentially methylated regions/bases based on cutoffs  
      methylDiff.obj = getMethylDiff(meth.diffmeth, 
                               difference=difference,
                               type="all",
                               qvalue=qvalue)
      # Get hypo-methylated
      methylDiff.obj.hypo = getMethylDiff(meth.diffmeth,
                                    difference=difference,
                                    type="hypo",
                                    qvalue=qvalue)
      # Get hyper-methylated
      methylDiff.obj.hyper = getMethylDiff(meth.diffmeth,
                                     difference=difference,
                                     type="hyper",
                                     qvalue=qvalue)

}else{
    create.empty.methylDiff = function(sampleids, assembly, context, treatment){
      new("methylDiff",
          sample.ids = sampleids, 
          assembly = assembly,
          context = context,
          treatment=treatment,
          # strand.aware is set to FALSe by defualt in methylKit@1.3.3
          destranded=FALSE, 
          resolution="base"
      )
    }
    methylDiff.obj = create.empty.methylDiff(meth.unite@sample.ids, assembly, context, treatment)
    methylDiff.obj.hypo = create.empty.methylDiff(meth.unite@sample.ids, assembly, context, treatment)
    methylDiff.obj.hyper = create.empty.methylDiff(meth.unite@sample.ids, assembly, context, treatment)

}



# Check if there are some differentially methylated cytosines
methylDiff.nonempty = nrow(methylDiff.obj)>1
if(!methylDiff.nonempty) print("There are no differentially methylated bases")
meth.unite.or.methylDiff.nonempty = meth.unite.nonempty | methylDiff.nonempty
```

```{r, show differential methylation, echo=TRUE, warning=FALSE, eval=methylDiff.nonempty}

# Show differentially methylated regions in a table.

methylDiff2df_roundint = function(methylDiff.obj, roundval=4){
  df = as(methylDiff.obj, "data.frame")
  df[,5:7] = round(df[,5:7], roundval)
  return(df)
}

## show all differentially methylated regions
datatable(methylDiff2df_roundint(methylDiff.obj),
              extensions = 'Buttons',
              options = list(
                dom = 'Bfrtip',
                buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                scrollX = TRUE),
              caption = "Differentially methylated bases",
              rownames = FALSE)
## show hyper-methylated region
datatable(methylDiff2df_roundint(methylDiff.obj.hyper),
              extensions = 'Buttons',
              options = list(
                dom = 'Bfrtip',
                buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                scrollX = TRUE),
              caption = "Hyper-methylated bases",
              rownames = FALSE)
## show hypo-methylated region
datatable(methylDiff2df_roundint(methylDiff.obj.hypo),
              extensions = 'Buttons',
              options = list(
                dom = 'Bfrtip',
                buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                scrollX = TRUE),
              caption = "Hypo-methylated bases",
              rownames = FALSE)

```


### Export differentially methylated cytosines

Finally we export the regions to a *BED* file, which can be loaded into any genome browser like [IGV](http://software.broadinstitute.org/software/igv/) or [UCSC](https://genome.ucsc.edu/) to allow for further analysis, annotation and visualisation. 

```{r export to bed, warning=FALSE}

#' Export windows to a BED file
#' 
#' The windows are color coded based on their score (methylation or differential
#' methylation value).
#' 
#' @param windows \code{\link[GenomicRanges]{GRanges}} object with information about
#' differentially methylated regions
#' @param filename name of the output data
#' @param trackLine UCSC browser trackline
#' @param colramp color scale to be used in the BED display
#' defaults to gray,green, darkgreen scale.
#' 
#' @return A BED files with the differentially methylated regions
#' which can be visualized in the UCSC browser 
#' 
#' @seealso \code{\link{methylKit::methSeg2bed}}
#' 
#' @export
#' @docType methods
#' @rdname meth2bed
meth2bed<-function(windows,filename,
    trackLine="track name='meth windows' description='meth windows' itemRgb=On",
    colramp=colorRamp(c("gray","green", "darkgreen"))
                        ){
  
  range01 <- function(x){(x-min(x)+1)/(max(x)-min(x)+1)}

  if(class(windows)!="GRanges"){
    windows=as(windows, "GRanges")
  }
  
  ## case if only one line is exported
  if(is.null(colramp) | length(windows)==1){
    trackLine <- gsub(pattern = "itemRgb=On",replacement = "",x = trackLine)
  } else {
    require(rtracklayer)
    ramp <- colramp
    score(windows)=range01(windows$meth.diff)
    mcols(windows)$itemRgb= rgb(ramp(score(windows)), maxColorValue = 255)     
  }
  
  strand(windows)="*"
  score(windows)=range01(windows$meth.diff)
  
  if(is.null(trackLine)){
    
    export.bed(windows,filename)
  }else{
    export.bed(windows,filename,
               trackLine=as(trackLine, "BasicTrackLine"))
  }
}


# Export differentially methylated bases to a bed file
trackLine = paste0("track name='differentially methylated bases ' ",
                         "description='diff. meth. between ",
                         paste(meth.diffmeth@sample.ids,collapse=","),
                         " mapped to ",
                         meth.diffmeth@assembly,
                         "' itemRgb=On")
meth2bed(windows = methylDiff.obj,
         trackLine=trackLine,
         colramp=colorRamp(c("gray","green", "darkgreen")),
         filename = output) 

```

The methylDiff object from the methylKit package is saved to enable fast loading for further analysis.

```{r}

# Save output of differential methylation calling into a RDS files
saveRDS(methylDiff.obj, methylDiff_file)
saveRDS(methylDiff.obj.hypo, methylDiff_hypo_file)
saveRDS(methylDiff.obj.hyper, methylDiff_hyper_file)


```


`r if(makeSessioninfo) {"### Session Info"}`

```{r ,include=makeSessioninfo  }
sessionInfo()
```


<!-- this section is used to generate a merged sessioninfo from multiple reports -->
```{r, include=FALSE }
session_file <- tempfile(pattern = 'session',tmpdir = ".",fileext = '.rds')
sess <- sessionInfo()
saveRDS(sess,file = session_file)
```


`r if(makeReferences) {"### References"}`