---
title: "PiGx: BS-seq"
output: 
  html_notebook:
    toc:        TRUE
    toc_float:  TRUE
    theme:      "lumen"
    number_sections: FALSE
    code_folding: "hide"
    self_contained: TRUE
    includes:
      in_header: pigx_bsseq_logo.html
date: "`r Sys.Date()`"
bibliography: reports.bib
# biblio-style: apalike
link-citations: yes
#params:
#  qc_path: "/Users/agosdsc/Development/Snakemake/git/pigx_bsseq/tmp/out2/01_rawqc/output2K_1_fastqc.html"
---

<!-- ```{r eval_params, echo=FALSE} -->

<!-- qc_path     <- params$qc_path -->
<!-- ``` -->

# Final Report {-}
  
This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

For now, you have to install the development versions of **bookdown** from Github:

```{r eval=FALSE}
devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need to install XeLaTeX.

```{r include=FALSE}
# automatically create a bib database for R packages
# knitr::write_bib(c(
# .packages(), 'bookdown', 'knitr', 'rmarkdown'
# ), 'packages.bib')
```

<!-- This setion needs to be updated to include links to the qc-reports or maybe to multiqc-report -->

<!-- ## Quality Control {-} -->

<!-- [qc_path](`r qc_path`) -->


