---
title: "PiGx: BS-seq"
date: '`r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`'
bibliography: reports.bib
params:
  rdata: ''
  grdata: ''
  outBed: ''
  sessioninfo: TRUE
  references: TRUE
---

```{r include=FALSE}
makeSessioninfo=params$sessioninfo
makeReferences = params$references
```

```{r eval_params, echo=FALSE}
input     <- params$rdata
output    <- params$outBed
grFile    <- params$grdata    
```

## Segmentation

The segmentation based on methylation provides a way to compress the information from base-pair into regional resolution, allowing the detection of regions with similar methylation values, that might be interesting for further investigations.

```{r, include=FALSE,warning=FALSE}

  suppressPackageStartupMessages(expr = {
  ## load methylKit
    library("methylKit",warn.conflicts = FALSE,quietly = TRUE)
  ## load DT 
    library("DT")
  })
```

### Libraries
```{r load libraries, eval=FALSE, include=TRUE}

  ## load methylKit
    library("methylKit")
  ## load DT 
    library("DT")

```

### Parameters
```{r print_params, echo=TRUE, results="asis"}

paramsDesc <- c("Input Data",
            "Output file")

paramsValues <- c( input,
             output)

inputParams <- data.frame(Parameters = paramsDesc,
                          Values = paramsValues,
                          stringsAsFactors = FALSE)
#params
#datatable(inputParams,options = list(dom = 't'),rownames = FALSE)
knitr::kable(inputParams)
```

### Segmentation of methylation profile

The segmentation of the methylation profile is done using the methSeg() function, where change-points in the genome wide  signal are recorded and the genome is partitioned into regions between consecutive change points. This approach is typically used in the context of copy number variation detection [@klambauer2012] but can be applied to methylome segmentation as well [@Wreczycka2017]. Here the identified segments are further clustered based on their average methylation signal using a mixture modeling approach allowing the detection of distinct regions inside the genome  [@Wreczycka2017].

```{r load and segment}

    load(input)

    ## Segmentation
    res.gr = methSeg(methRaw,diagnostic.plot=TRUE)

    ## convert to data.frame
    res.df <- as.data.frame(res.gr)
    res.df$seg.mean <- round(res.df$seg.mean,digits = 3)

    ## show the regions
    datatable(res.df,
              extensions = 'Buttons', 
              options = list(
                dom = 'Bfrtip',
                buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                scrollX = TRUE),
              caption = "Resulting Segments",
              rownames = FALSE)
```

The GRanges object is saved to enable fast loading for further analysis.
```{r}
    ## Saving object
    save(res.gr,file=grFile) 
```



### Export

Finally we export the regions to a *BED* file, which can be loaded into any genome browser like [IGV](http://software.broadinstitute.org/software/igv/) or [UCSC](https://genome.ucsc.edu/) to allow for further analysis, annotation and visualisation. 
```{r export to bed}
    ## export segments to bed file
    methSeg2bed(segments = res.gr,
                trackLine = paste0("track name='meth segments ' ",
                                   "description='meth segments of ",
                                    methRaw@sample.id,
                                   " mapped to ",
                                   methRaw@assembly,
                                   "' itemRgb=On"),
                colramp=colorRamp(c("gray","green", "darkgreen")),
                filename = output)

```


`r if(makeSessioninfo) {"### Session Info"}`

```{r ,include=makeSessioninfo  }
sessionInfo()
```


<!-- this section is used to generate a merged sessioninfo from multiple reports -->
```{r, include=FALSE }
session_file <- tempfile(pattern = 'session',tmpdir = ".",fileext = '.rds')
sess <- sessionInfo()
saveRDS(sess,file = session_file)
```


`r if(makeReferences) {"### References"}`
