---
title: "CRISPR Report"
output: html_document
date: '`r date()`'
params: 
  sampleSheetFile: ''
  ampliconFastaFile: ''
  ampliconName: ''
  cutSitesFile: ''
  indelsFolder: ''
  workdir: ''
  prefix: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = params$workdir)
library(knitr)
library(ggplot2)
library(ggrepel)
library(data.table)
library(rtracklayer)
library(DT)
library(plotly)
library(crosstalk)
library(GenomicAlignments)
```

# Input Settings
```{r printInputSettings}
ampliconFastaFile <- params$ampliconFastaFile
ampliconName <- params$ampliconName
sampleSheetFile <- params$sampleSheetFile
indelsFolder <- params$indelsFolder
cutSitesFile <- params$cutSitesFile
prefix <- params$prefix
workdir <- params$workdir

inputParameterDesc <- c('Amplicon Fasta File',
                        'Amplicon Name',
                        'sgRNA Cut Sites File', 
                     'Experiment Data File',
                     'indels folder path', 
                     'Prefix for output files',
                     'Working directory'
                     )
inputParameterValues <- c(ampliconFastaFile,
                          ampliconName,
                          cutSitesFile, 
                          sampleSheetFile,
                          indelsFolder, 
                          prefix,
                          workdir)
inputSettings <- data.frame(parameters = inputParameterDesc,
                            values = inputParameterValues,
                            stringsAsFactors = FALSE)
DT::datatable(data = inputSettings,
              extensions = 'FixedColumns',
              options = list(fixedColumns = TRUE,
                         scrollX = TRUE,
                         pageLength = nrow(inputSettings),
                         dom = 't'))
```

```{r printSampleInformation}
sampleSheet <- read.csv(sampleSheetFile, stringsAsFactors = F)
sampleSheet <- sampleSheet[sampleSheet$amplicon == ampliconName,]
DT::datatable(data = sampleSheet,
              extensions = 'FixedColumns',
              options = list(fixedColumns = TRUE,
                         scrollX = TRUE,
                         dom = 't'))

cutSites <- read.table(cutSitesFile, stringsAsFactors = F)
colnames(cutSites) <- c('sgRNA', 'cutSite')
DT::datatable(data = cutSites)

```

# Coverage Profiles 

## Coverage as Heatmap
```{r coverage_heatmap}
# import stats tables 
coverageStats <- as.data.table(do.call(rbind, lapply(sampleSheet$sample_name, function(sampleName) {
  f <- file.path(indelsFolder, paste0(sampleName, '.coverageStats.tsv'))
  if(file.exists(f)) {
    dt <- data.table::fread(f)
    return(dt)
  }
})))
#create a matrix of bp versus samples where values are coverage
dt <- dcast.data.table(coverageStats, sample ~ bp, value.var = 'cov', fill = 0)
M <- as.matrix(dt[,-1])
rownames(M) <- dt$sample
pheatmap::pheatmap(M, 
                   cluster_rows = nrow(M) > 1, 
                   cluster_cols = F, 
                   show_colnames = F, 
                   main = paste(ampliconName, 'Coverage Profiles'))
```

## Coverage as Line Plot

```{r coverage-lineplot}
p <- ggplot(coverageStats, aes(x = bp, y = cov, group = sample), height = 500) + 
  geom_line(aes(color = sample), show.legend = F) +   
  theme_bw() +  
  labs(x = 'base position', y = 'Number of Reads', title = paste(ampliconName, 'Coverage Profiles'))

ggplotly(p)
```

# Indel Score Profiles 

```{r plotIndelProfiles, fig.height=8}

plotScores <- function(indelsFolder, sampleName, sampleSheet, cutSites) {
  f <- file.path(indelsFolder, paste0(sampleName, '.coverageStats.tsv'))
  dt <- data.table::fread(f)
  dt$coverage <- dt$cov / max(dt$cov) #add relative coverage score 

  sgRNAs <- unlist(strsplit(x = sampleSheet[sampleSheet$sample_name == sampleName,]$sgRNA_ids, 
                     split = ':'))
  
  sgRNAs <- merge(data.frame('sgRNA' = sgRNAs, 'sample_name' = sampleName), cutSites, by = 'sgRNA')

  mdt <- melt(dt[,-c('ins', 'del', 'indel', 'cov')], id.vars = c('bp', 'sample', 'seqname'))
  
  p <- ggplot(mdt, aes(x = bp, y = value, group = variable)) +
    geom_line(aes(color = variable)) + labs(title = sampleName) 
  if(nrow(sgRNAs) > 0) {
    p <- p + geom_vline(data = sgRNAs, aes(xintercept = cutSite, 
                                           color = sgRNA), linetype = 'dotted')
  }
  p <- p + 
    scale_y_continuous(labels = scales::percent, 
                       breaks = seq(0, 1, by = 0.025), 
                       limits = c(0, 1))
  return(p)
}

plots <- lapply(sampleSheet$sample_name, function(sampleName) {
  plotScores(indelsFolder = indelsFolder, 
             sampleName = sampleName, 
             sampleSheet = sampleSheet, 
             cutSites = cutSites)
})
names(plots) <- sampleSheet$sample_name
```

```{r plotIndelProfilesHeader, results = 'asis'}
if(length(plots) > 10) {
  cat("## Indel Score Profiles {.tabset .tabset-dropdown}\n\n")
} else {
  cat("## Indel Score Profiles {.tabset}\n\n")
}
```

```{r plotIndelProfilesPlots}
out = NULL
for (n in names(plots)) {
  p <- ggplotly(plots[[n]])
  out = c(out, knitr::knit_expand(text='### {{n}} \n\n {{p}} \n\n'))
}
```

`r paste(knit(text = out), collapse = '\n')`

# sgRNA efficiencies at cut sites
```{r cutSiteStats}
# this is a table for all samples versus all cut sites in the amplicon
# not all sample x sgRNA combinations are true, 
# but still they are also computed to serve as a negative control
cutSiteStats <- as.data.table(do.call(rbind, lapply(sampleSheet$sample_name, function(sampleName) {
  f <- file.path(indelsFolder, paste0(sampleName, '.indel_stats_at_cutsites.tsv'))
  if(file.exists(f)) {
    dt <- fread(f)
    return(dt)
  }
})))

# match samples to the actual sgRNA guides used in that sample
sampleGuides <- lapply(sampleSheet$sample_name, function(s) {
  sgRNAs <- unlist(strsplit(x = sampleSheet[sampleSheet$sample_name == s,]$sgRNA_ids, 
                   split = ':'))
})
names(sampleGuides) <- as.character(sampleSheet$sample_name)

cutSiteStats$sampleMatchesGuide <- as.factor(apply(cutSiteStats, 1, function(x) {
  s <- as.character(x[['sample']])
  g <- as.character(x[['sgRNA']])
  return(g %in% sampleGuides[[s]])
}))

plots <- lapply(as.vector(unique(cutSiteStats$sgRNA)), function(sg) {
  dt <- cutSiteStats[sgRNA == sg & sampleMatchesGuide == TRUE]
  if(nrow(dt) > 0) {
    p <- ggplot(dt, aes(x = sample, y = indelEfficiency)) + 
      geom_bar(stat = 'identity', aes(fill = sample)) + ylim(0, 100) + 
      labs(title = sg) + coord_flip()
    return(p)
  }
})
names(plots) <- as.vector(unique(cutSiteStats$sgRNA))
```


```{r plotIndelStatsHeader, results = 'asis'}
if(length(plots) > 10) {
  cat("## Indel Stats At Cut Sites {.tabset .tabset-dropdown}\n\n")
} else {
  cat("## Indel Stats At Cut Sites {.tabset}\n\n")
}
```

```{r plotIndelStatsPlots}
out = NULL
for (n in names(plots)) {
  if(!is.null(plots[[n]])) {
  p <- ggplotly(plots[[n]])
  out = c(out, knitr::knit_expand(text='### {{n}} \n\n {{p}} \n\n'))
  } else {
    out = c(out, knitr::knit_expand(text='### {{n}} \n\n No indel data found\n\n'))
  }
}
```

`r paste(knit(text = out), collapse = '\n')`

# sgRNA efficiencies at cut sites - version 2

```{r}
dts <- crosstalk::SharedData$new(cutSiteStats)

bscols(widths = c(4,8),
  list(
    filter_checkbox("sampleMatchesGuide", "Own Guide", dts, ~factor(sampleMatchesGuide), inline = TRUE),
    filter_select( "sgRNA", "Select sgRNA", dts, ~factor(sgRNA), allLevels = FALSE, multiple = FALSE),
    filter_select( "sample", "Select samples", dts, ~factor(sample), allLevels = FALSE, multiple = TRUE)
    ),
  plotly::plot_ly(data = dts,
                  x = ~ sgRNA,
                  y = ~ indelEfficiency,
                  group = ~sgRNA,
                  color = ~sample,
                  type = 'bar',
                  text = ~sample,
                  height = 500)  %>%
    layout(showlegend = FALSE, xaxis = list(showticklabels = TRUE), barmode = 'group')
)
```

# Indel diversity at cutsites {.tabset}

```{r}
# indels: a data.table object with minimal columns: start, end, 
# cutSites: a data.frame/data.table object with minimal columns: sgRNA and cutsite
# return: data.frame (nrow = nrow(indels), columns are sgRNA ids, 
#         values are 1 if indel overlaps cutsite, otherwise 0. 
overlapCutSites <- function(indels, cutSites, extend = 3) {
  target <- IRanges(start = cutSites$cutSite - extend, 
                    end = cutSites$cutSite + extend, 
                    names = cutSites$sgRNA)
  query <- IRanges(start = indels$start, end = indels$end)
  
  startOverlaps <- as.data.table(findOverlaps(start(query), target, type = 'any'))
  endOverlaps <- as.data.table(findOverlaps(end(query), target, type = 'any'))
  
  overlaps <- merge(startOverlaps, endOverlaps, by = c('queryHits', 'subjectHits'), all = TRUE)
  
  M <- matrix(data = rep(0, nrow(indels) * length(target)), 
              nrow = nrow(indels), ncol = length(target))
  colnames(M) <- names(target)

  M[as.matrix(overlaps)] <- 1
  
  return(M)
}

#import all detected indels (unfiltered) from all samples
allIndels <- as.data.table(do.call(rbind, lapply(sampleSheet$sample_name, function(sampleName) {
  f <- file.path(indelsFolder, paste0(sampleName, '.indels.unfiltered.tsv'))
  if(file.exists(f)) {
    dt <- fread(f)
    return(dt)
  }
})))

#summarize indels by read support
indelSummary <- allIndels[,length(unique(readID)), by = c('seqname', 'sample', 'start', 'end', 'indelType')]
colnames(indelSummary)[6] <- 'ReadSupport'

dt <- cbind(indelSummary, 
            overlapCutSites(indels = indelSummary, 
                            cutSites = cutSites, extend = 3))

mdt <- melt(dt, id.vars = colnames(indelSummary))

#make plots for different score thresholds, facet by samples, guides, and indel types
supportThresholds <- c(0, 1, 5, 10)
plots <- lapply(supportThresholds, function(s) {
  df <- mdt[ReadSupport > s & value == 1,
            length(seqname), 
            by = c('sample', 'variable', 'indelType')]
  if(nrow(df) > 0) {
    p <- ggplot(data = df, aes(x = sample, y = variable)) + 
    #geom_point(aes(size = V1)) + 
    geom_tile(aes(fill = V1)) 
    if(nrow(df) < 250) {
      p <- p + 
        geom_text(aes(label = V1), color = 'white') + facet_wrap(~ indelType, ncol = 1) 
    }
    p <- p + theme_bw() + 
    theme(axis.text.x = element_text(angle = 90), legend.position = 'none')
    return(p)
  } else {
    NULL
  }
})
names(plots) <- paste("Read Support >",supportThresholds)
```

```{r plotGenotypeDiversity, results='asis', fig.height=8}
for (i in 1:length(plots)) {
  cat("## ",names(plots)[i],"\n\n")
  p <- plots[[i]]
  if(!is.null(p)) {
    print(plots[[i]])
  } else {
    cat("No genotypes detected")
  }
  cat("\n\n")
}
```

# Diversity of Deletions {.tabset}

Here we plot the locations and diversity of the deletions that originate at the cut sites (+/- 3 bps) and are supported by at least 5 reads. 

```{r deletionDiversity}

deletionCutsiteOverlaps <- cbind(indelSummary[indelType == 'D'], overlapCutSites(indelSummary[indelType == 'D'], cutSites))

plots <- lapply(unique(deletionCutsiteOverlaps$sample), function(s) {
  dt <- deletionCutsiteOverlaps[sample == s & ReadSupport > 5]
  if(nrow(dt) == 0) {
    return(NULL)
  }
  sgRNAs <- unlist(strsplit(x = sampleSheet[sampleSheet$sample_name == s,]$sgRNA_ids, 
                            split = ':'))
  if(sgRNAs[1] == 'none') {
    return(NULL)
  }
  dt <- subset(dt, select = c('seqname', 'start', 'end', 'indelType', 'ReadSupport', sgRNAs))
  # select indels that overlap at least one cut site
  dt <- dt[apply(subset(dt, select = sgRNAs), 1, function(x) sum(x > 0) > 0),]
  
  dt <- dt[order(end - start)]
  dt$linePos <- 1:nrow(dt)
  p1 <- ggplot2::ggplot(dt, aes(x = linePos, ymin = start, ymax = end)) + 
    geom_linerange(size = 0.5) + 
    labs(title = paste0('unique deletions with > 5 reads at cut sites: n=',nrow(dt))) + 
    geom_point(data = dt, aes(x = linePos, y = start), size = 1, color = 'red') + 
    geom_point(data = dt, aes(x = linePos, y = end), size = 1, color = 'blue') + 
    geom_hline(data = cutSites[cutSites$sgRNA %in% sgRNAs,], 
              aes(yintercept = cutSite, color = sgRNA), show.legend = FALSE) +
    theme(axis.text.y = element_blank(), 
          axis.title.y = element_blank(), 
          axis.title.x = element_blank(),
          axis.ticks.y = element_blank(), 
          axis.text.x = element_text(angle = 90), 
          plot.title = element_text(hjust = 0.5)) + 
    scale_y_continuous(sec.axis = dup_axis(breaks = cutSites[cutSites$sgRNA %in% sgRNAs,]$cutSite,
                                           labels = cutSites[cutSites$sgRNA %in% sgRNAs,]$sgRNA)) + 
    coord_flip()
  
  p2 <- ggplot(dt, aes(x = 1, y = end - start + 1)) + 
          geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
          geom_jitter(height = 0, alpha = 0.5, 
                      aes(size = as.factor(round(log10(dt$ReadSupport))), 
                          color = as.factor(round(log10(dt$ReadSupport))))) + 
          labs(x = '', y = 'Deletion Length', 
               size = 'log10(Read Support)', 
               color = 'log10(Read Support)') + 
          theme(axis.ticks.x = element_blank() , axis.text.x = element_blank(), 
                legend.position = 'bottom')

  p <- cowplot::plot_grid(p1, p2, labels = c('A', 'B'))
  return(p)
})
names(plots) <- unique(deletionCutsiteOverlaps$sample)
```

```{r plotDeletionDiversity, results='asis', fig.height=8, fig.width=10}
for (n in names(plots)) {
  cat('##',n,'\n\n')
  p <- plots[[n]]
  if(!is.null(p)) {
        print(p)
    } else {
        cat("No plot to show\n\n")
      }
  cat("\n\n")
}
```

# R Session Information
```{r}
print(sessionInfo())
```

